Tracking $p$-adic precision

### 2014-02-24 padicprec.tex
Small corrections.

The approach of propagating precision with each arithmetic operation works well, but does
sometimes underestimate the known precision of a result, as we will discuss in Section
\ref{ssec:separation}.  Moreover, elements of $\Qp$ provide building blocks for generic
implementations of polynomials, vector spaces, matrices and power series.  The practice
of storing the precision within each entry is not flexible enough for all applications.
Sometimes only a rough accounting of precision is needed, in which case storing
and computing the precision of each entry in a large matrix needlessly consumes
space and time.  Conversely, recording the precision of each entry does not allow
a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
quadratic polynomial $f$.

@@

\subsection{Problems in precision}

We begin by offering two examples that illustrate key features of $p$-adic precision.
Recall that an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if
it is well defined up to adding an unknown element of $p^N\Zp$.
The following problems illustrate cases where step-by-step precision tracking
does not yield optimal results.

@@

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.

Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric -- 
\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
-- and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
We underline that any lattice $H$ in $E$ is also closed since its 
complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
are all open.

For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
\begin{align*}
B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
\end{align*}
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written
\[
x = \sum_{i \in I} \alpha_i x_i
\]
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

@@

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
\begin{deftn}
\begin{itemize}
\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}

@@

The definition of an approximate element encapsulates the two main practical suggestions of this paper
with regards to representing vector spaces, matrices, polynomials and power series over $K$:
\begin{enumerate}
\item one should \textbf{separate} the approximation from the precision,
\item the appropriate object to represent precision is a \textbf{lattice}.
\end{enumerate}
In the rest of this section we discuss some of the benefits made possible these choices.

Note first that using an arbitrary lattice to represent the precision of an approximate element
can reduce precision loss when compared to storing the precision of each coefficient
$\alpha_i$ in \eqref{eq:exact_elt} separately.  Recall the map 
$f : (x,y) \mapsto (x+y, x-y)$ from the beginning of the section, and write $(e_1, e_2)$
for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
By using lattices one eliminates the loss of precision seen previously.

@@

\begin{deftn} \label{deftn : diff}
Let $E$ and $F$ be two normed $K$-vector spaces, let $U$ be an open 
subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
there exists an open neighborhood $U_\varepsilon \subset U$ containing 
$v_0$ with
\[ 
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

One can derive all usual properties of the derivative from this 
definition: unicity, chain rules, \emph{etc}.

@@

Then, for all $\rho \in (0, 1]$, there exists a positive real 
number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
$H$ such that $B^-_E(\rho r) \subset H \subset B_E(r)$ is a first
order lattice for $f$ at $v_0$.
\end{lem}

@@

Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with 
$B^-_E(\rho r) \subset H \subset B_E(r).$
We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
that $f(H) \subset f'(0) (H)$. Suppose $x \in H$. By differentiability at 
$0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
Setting $y=f(x)-f'(0)(x)$, we have $\Vert y \Vert \leq 
\varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
\subset f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
\in B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
$\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
then $f(x)= f'(0) (x-x') \in f'(0) (H)$.

We shall now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
be such that $y = f'(0) (x_0)$. We inductively define two sequences 
$(x_n)$ and $(z_n)$ as follows:
\begin{itemize}
\item $z_n$ is an element of $E$ satisfying $f'(0)(z_n) = y - 
f(x_n)$ and $\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$ (such 
an element exists by definition of $C$), and
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
For convenience, let us also define $x_{-1} = 0$ and $z_{-1}=x_0.$ We claim that the 
sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
$H$. We do so by induction, assuming that $x_{n-1}$ and $x_n$ belong to $H$
and showing that $z_n$ and $x_{n+1}$ do as well. Noticing
that 
\begin{equation}
\label{eq:mainlemma}
\begin{aligned}
y - f(x_n) &= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n) \\
&= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
\end{aligned}
\end{equation}
we deduce using differentiability that
$\Vert y - f(x_n) \Vert \leq \varepsilon \cdot \Vert x_n - x_{n-1}
\Vert$.
Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
deduce $x_{n+1} \in H$.

@@

For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
in a neighborhood of $0$. Under this assumption, the datum of $f$ 
uniquely determines the $f_n$'s (a consequence of Proposition 
\ref{prop:lambdaNP}).
To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
\cup \{+\infty\}$ defined by:
$$\begin{array}{rcll}
\Lambda(f)(v) & = & 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \big)
& \text{if } f \text{ is defined on } B^-_E(e^v) \\
& = & +\infty & \text{otherwise}
\end{array}$$
The following lemma is easy and left to the reader.

@@

\begin{proof}
Let $v \in \R$. We assume first that $\NP(f)^\star(v)$ is finite and 
that $v$ is not a slope of $\NP(f)$. We set $u = \NP(f)^*(v)$. The 
function $m \mapsto \NP(f)(m) - vm + u$ has the following properties:
\begin{enumerate}
\item it is piecewise affine and everywhere nonnegative,
\item it does not admit $0$ as a slope and
\item it vanishes at $x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
\end{enumerate}
We deduce from these facts that there exists $c>0$ such that 
\[
vm-u \leq - \log \lVert f_m \rVert - c \cdot \vert n-m \vert
\]
for any $m\geq 0$.
Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get 
\[
-vn-\log \lVert f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert.
\]
Therefore, for any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
\[
\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.
\]
Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get:
\begin{equation}
\label{eq:lowerLambda}
\Lambda (f)(v) \leq \log \big( \Vert f_n \Vert e^{vn} \big)
= vn + \log \Vert f_n \Vert = u.
\end{equation}
On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
and the fact that $\vert K^\star \vert$ is dense in $\R$ (which holds
because $K$ is algebraically closed)
that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
$\lim_{i \to \infty} \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
furthermore $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} 
\Vert f_n \Vert \cdot e^{vn}$ for all $i$, we get $\Vert f_m (x_i) \Vert 
< \Vert f_n (x_i) \Vert$ for $i$ large enough. For these $i$, we then 
have $\Vert f(x_i) \Vert = \Vert f_n (x_i) \Vert$. Passing to the limit 
on $i$, we find $\Lambda (f)(v) \geq u$. Comparing with 
\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
\NP(f)^\star(v)$.

@@

We now assume that $\NP(f)^\star(v) = +\infty$. The function $x \mapsto 
\NP(f)(x) - vx$ is then not bounded from below. Since it is convex, it 
goes to $-\infty$ when $x$ goes to $+\infty$. Going back to the 
definition of $\NP(f)$, this implies that $v n + \log \Vert f_n \Vert$
goes to infinity when $n$ grows. It is then enough to establish the
following claim:
\begin{equation}
\label{eq:claim}
\forall n \in \N, \quad 
\Lambda(f)(v) \geq v n + \log \Vert f_n \Vert - \log 2.
\end{equation}
Let $n$ be a fixed integer. If $\Vert f_n 
\Vert = 0$, there is nothing to prove. Otherwise, let us consider an
element $x_n \in B^-_E(e^v)$ such that $\Vert f_n(x_n) \Vert \geq \frac 
1 2 \Vert f_n \Vert \cdot e^{vn}$. If the series $\sum_{m \geq 0} 
f_m(x_n)$ diverges, then $\Lambda(f)(v) = +\infty$ by definition and 
Eq.~\eqref{eq:claim} holds. On the other hand, if it converges, the
sequence $\Vert f_m(x_n) \Vert$ goes to $0$ when $m$ goes to infinity.
Hence it takes its maximum value $R$ a finite number of times; let us 
denote by $I \subset \N$ the set of the corresponding indices. For any
$\lambda \in B_K(1)$, the series defining $f(\lambda x_n)$ converges 
and
$$f(\lambda x_n) \in B_F(R)
\quad \text{and} \quad
f(\lambda x_n) \equiv \sum_{m \in I} \lambda^m f_m(x_n)
\pmod {B^-_F(R)}.$$
The quotient $B_F(R) / B^-_F(R)$ is a vector space over the residue
field of $K$, that is the field $k$ defined by $k = B_K(1)/B^-_K(1)$.
Hensel's lemma implies that $k$ is algebraically closed (because $K$ 
is) and then infinite. Therefore, there must exist $\lambda \in 
B_K(1)$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
in $B_F(R) / B^-_F(R)$. For such an element $\lambda$, we have $\Vert 
f(\lambda x_n) \Vert = R \geq \frac 1 2 \Vert f_n \Vert \cdot e^{vn}$. 
The claim \eqref{eq:claim} follows.

@@

We have then proved that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
coincide expect maybe on the set consisting of slopes of $\NP(f)$. But 
we remark that this set is countable and that both functions $\Lambda 
(f)$ and $\NP(f)^\star$ are left continuous. This is enough to ensure 
that $\Lambda (f)$ and $\NP(f)^\star$ coincide everywhere.
\end{proof}

@@

\begin{rem}
On a neighborhood of $-\infty$, the function $x \mapsto \Lambda(f)_{\geq 
2} (x) - x$ is affine of slope $1$. This implies that, for all $\rho \in
(0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
if $\rho$ is close enough to $0$, then one can take $\delta = e^\nu$ as a 
linear function of $\rho$.
\end{rem}

@@

We consider a function {\tt f} (in the sense of computer 
science) that takes as input an approximate element lying in an open 
subset $U$ of a $K$-Banach space $E$ and outputs another approximate
element lying in an open subset $V$ of another $K$-Banach space $F$.
In applications, this function models a continuous
mathematical function $f : U \to V$: when 
{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
$f(x+H) \subseteq x' + H'$. We say that {\tt f} \emph{preserves precision}
if the above inclusion is an equality; it is 
often not the case as shown in Section \ref{ssec:separation}.

>>>>

The approach of propagating precision with each arithmetic operation works well, but does
sometimes underestimate the known precision of a result, as we will discuss in Section
\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
implementations of polynomials, vector spaces, matrices and power series.  The practice
of storing the precision within each entry is not flexible enough for all applications.
Sometimes only a rough accounting of precision is needed, in which case storing
and computing the precision of each entry in a large matrix needlessly consumes
space and time.  Conversely, recording the precision of each entry does not allow
a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
quadratic polynomial $f$.

@@

\subsection{Problems in precision}
\label{ssec:stepbystep}

The usual way to track $p$-adic precision consists in replacing $p$-adic 
numbers by approximate elements of the form $a + O(p^N)$ and performing 
all usual arithmetical operations on these approximations.
%Recall that 
%an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if it 
%is well defined up to adding an unknown element of $p^N\Zp$.
We offer below three examples that illustrate cases where this way to 
track precision does not yield optimal results.

@@

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
We underline that any lattice $H$ in $E$ is also closed since its 
complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
are all open.
For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

@@

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
\begin{deftn}
\label{def:approximate}
\begin{itemize}
\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}

@@

Definition \ref{def:approximate} encapsulates the two main practical suggestions of this paper
with regards to representing vector spaces, matrices, polynomials and power series over $K$:
\begin{enumerate}
\item one should \textbf{separate} the approximation from the precision,
\item the appropriate object to represent precision is a \textbf{lattice}.
\end{enumerate}
In the rest of this section we discuss some of the benefits made possible these choices.

Note first that using an arbitrary lattice to represent the precision of an approximate element
can reduce precision loss when compared to storing the precision of each coefficient
$\alpha_i$ in \eqref{eq:exact_elt} separately.  Recall the map 
$f : (x,y) \mapsto (x+y, x-y)$ from the beginning of the section, and write $(e_1, e_2)$
for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
By using lattices one eliminates the loss of precision seen previously.
We shall see in the next section that a similar phenomenon occurs for non-linear mappings as well.

@@

\begin{deftn} \label{deftn : diff}
Let $E$ and $F$ be two $K$-Banach spaces, let $U$ be an open 
subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
there exists an open neighborhood $U_\varepsilon \subset U$ containing 
$v_0$ with
\[ 
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

%One can derive all usual properties of the derivative from this 
%definition: unicity, chain rules, \emph{etc}.

@@

Then, for all $\rho \in (0, 1]$, there exists a positive real 
number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ is a first
order lattice for $f$ at $v_0$.
\end{lem}

@@

Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with 
$B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r).$
We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
that $f(H) \subset f'(0) (H)$. Suppose $x \in H$. By differentiability at 
$0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
Setting $y=f(x)-f'(0)(x)$, we have $\Vert y \Vert \leq 
\varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
\subset f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
\in B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
$\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
then $f(x)= f'(0) (x-x') \in f'(0) (H)$.

We now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
be such that $y = f'(0) (x_0)$. We inductively define two sequences 
$(x_n)$ and $(z_n)$ as follows:
\begin{itemize}
\item $z_n$ is an element of $E$ satisfying $f'(0)(z_n) = y - 
f(x_n)$ and $\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$ (such 
an element exists by definition of $C$), and
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
For convenience, let us also define $x_{-1} = 0$ and $z_{-1}=x_0.$ We claim that the 
sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
$H$. We do so by induction, assuming that $x_{n-1}$ and $x_n$ belong to $H$
and showing that $z_n$ and $x_{n+1}$ do as well. Noticing
that 
\begin{equation}
\label{eq:mainlemma}
\begin{aligned}
y - f(x_n) &= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n) \\
&= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
\end{aligned}
\end{equation}
we deduce using differentiability that
$\Vert y - f(x_n) \Vert \leq \varepsilon \cdot \Vert x_n - x_{n-1}
\Vert$.
Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
deduce $x_{n+1} \in H$.

@@

For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
in a neighborhood of $0$. Under this assumption, the datum of $f$ 
uniquely determines the $f_n$'s (a consequence of Proposition 
\ref{prop:lambdaNP} below).
To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
\cup \{+\infty\}$ defined by:
$$\begin{array}{rcll}
\Lambda(f)(v) & = & 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \big)
& \text{if } f \text{ is defined on } B^-_E(e^v) \\
& = & +\infty & \text{otherwise}
\end{array}$$
The following lemma is easy and left to the reader.

@@

\begin{proof}
We first remark that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
are both left continuous. It is then enough to prove that they are
equal on a dense subset of $\R$. We are going to prove that they
coincide expect maybe on the set consisting of slopes of $\NP(f)$.

@@

Let $v \in \R$ which is not a slope of $\NP(f)$. We assume first that 
$\NP(f)^\star(v)$ is finite. We set $u = \NP(f)^*(v)$. The 
function $m \mapsto \NP(f)(m) - vm + u$ has the following properties:
\begin{enumerate}
\item it is piecewise affine and everywhere nonnegative,
\item it does not admit $0$ as a slope and
\item it vanishes at $x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
\end{enumerate}
We deduce from these facts that there exists $c>0$ such that 
\[
vm-u \leq - \log \lVert f_m \rVert - c \cdot \vert n-m \vert
\]
for any $m\geq 0$.
Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get 
\[
-vn-\log \lVert f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert.
\]
Therefore, for any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
\[
\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.
\]
Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get:
\begin{equation}
\label{eq:lowerLambda}
\Lambda (f)(v) \leq \log \big( \Vert f_n \Vert e^{vn} \big)
= vn + \log \Vert f_n \Vert = u.
\end{equation}
On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
and the fact that $\vert K^\star \vert$ is dense in $\R$ (which holds
because $K$ is algebraically closed)
that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
$\lim_{i \to \infty} \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
furthermore $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} 
\Vert f_n \Vert \cdot e^{vn}$ for all $i$, we get $\Vert f_m (x_i) \Vert 
< \Vert f_n (x_i) \Vert$ for $i$ large enough. For these $i$, we then 
have $\Vert f(x_i) \Vert = \Vert f_n (x_i) \Vert$. Passing to the limit 
on $i$, we find $\Lambda (f)(v) \geq u$. Comparing with 
\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
\NP(f)^\star(v)$.

@@

We now assume that $\NP(f)^\star(v) = +\infty$. The function $x \mapsto 
\NP(f)(x) - vx$ is then not bounded from below. Since it is convex, it 
goes to $-\infty$ when $x$ goes to $+\infty$. Going back to the 
definition of $\NP(f)$, this implies that $v n + \log \Vert f_n \Vert$
goes to infinity when $n$ grows. It is then enough to establish the
following claim:
\begin{equation}
\label{eq:claim}
\forall n \in \N, \quad 
\Lambda(f)(v) \geq v n + \log \Vert f_n \Vert - \log 2.
\end{equation}
Let $n$ be a fixed integer. If $\Vert f_n 
\Vert = 0$, there is nothing to prove. Otherwise, we consider an
element $x_n \in B^-_E(e^v)$ such that $\Vert f_n(x_n) \Vert \geq \frac 
1 2 \Vert f_n \Vert \cdot e^{vn}$. If the series $\sum_{m \geq 0} 
f_m(x_n)$ diverges, then $\Lambda(f)(v) = +\infty$ by definition and 
Eq.~\eqref{eq:claim} holds. On the other hand, if it converges, the
sequence $\Vert f_m(x_n) \Vert$ goes to $0$ when $m$ goes to infinity.
Hence it takes its maximum value $R$ a finite number of times; let us 
denote by $I \subset \N$ the set of the corresponding indices. For any
$\lambda \in B_K(1)$, the series defining $f(\lambda x_n)$ converges 
and
$$f(\lambda x_n) \in B_F(R)
\quad \text{and} \quad
f(\lambda x_n) \equiv \sum_{m \in I} \lambda^m f_m(x_n)
\pmod {B^-_F(R)}.$$
The quotient $B_F(R) / B^-_F(R)$ is a vector space over the residue
field of $K$, that is the field $k$ defined by $k = B_K(1)/B^-_K(1)$.
It is clearly algebraically closed (because $K$ 
is) and then infinite. Therefore, there must exist $\lambda \in 
B_K(1)$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
in $B_F(R) / B^-_F(R)$. For such an element $\lambda$, we have $\Vert 
f(\lambda x_n) \Vert = R \geq \frac 1 2 \Vert f_n \Vert \cdot e^{vn}$. 
The claim \eqref{eq:claim} follows.
\end{proof}

@@

\begin{rem}
On a neighborhood of $-\infty$, the function $x \mapsto \Lambda(f)_{\geq 
2} (x) - x$ is affine with slope $1$. This implies that, for all $\rho 
\in (0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
if $\rho$ is close enough to $0$, then one can take $\delta = e^\nu$ as a 
linear function of $\rho$.
\end{rem}

@@

We consider a function {\tt f} (in the sense of computer 
science) that takes as input an approximate element lying in an open 
subset $U$ of a $K$-Banach space $E$ and outputs another approximate
element lying in an open subset $V$ of another $K$-Banach space $F$.
In applications, this function models a continuous
mathematical function $f : U \to V$: when 
{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
$f(x+H) \subseteq x' + H'$. We say that {\tt f} \emph{preserves precision}
if the above inclusion is an equality; it is 
often not the case as shown in Section \ref{ssec:stepbystep}.
